---
title: "Tidycensus Tutorial"
author: "Myles Ndiritu"
output: pdf_document
date: "2023-09-26"
urlcolor: purple
---

Let's go over how to use Tidycensus to find and extract census data.

## STEP 0: Load the required packages:

```{r setup, warning=FALSE, message=FALSE}
library(tidycensus)
library(tidyverse)
```

## STEP 1: Run API key

You will need an API key to access the data. I have included my key below, but in the event that it doesn't work visit this website to get a key: <https://api.census.gov/data/key_signup.html>
  
```{r api, include=FALSE}
census_api_key("f82508cd291cb6c902d7e0558e5d8962e5900ba5")

```

```{r api1, eval=FALSE}
census_api_key("YOUR KEY HERE")

```

## STEP 2: Loading U.S. Decennial Census data

```{r d1}
v20 <- load_variables(2020, "pl", cache = T)
View(v20)
head(v20, 7)
```

Please become familiar with this format, as we will use it repeatedly when loading census data.

Adjust the year value in the code to change the year. 2020 is the most recent.

Examine the name column in this v20 data frame. Each code corresponds to a specific variable. For the next step, lets look at the black population variable, code name "P1_004N."

## STEP 3: Loading specific datasets

```{r d2}

black2020 <- get_decennial(geography = "state",
                             variables = "P1_004N",
                             year = 2020)
head(black2020)

```

The value column shows the black population for that county.

We'll also need to be familiar with the American Community Survey (ACS) data. Luckily, the method of extracting it is the same.

## STEP 4: Loading American Community Survey Data

Again, load the data set to see the variables and variable code names:

```{r d3}

acs20 <- load_variables(2020, "acs5", cache = T)
View(acs20)

```

## STEP 5: Loading specific ACS datasets

After finding a variable you want, extract it by defining the required parameters, and specifying the variable name:

```{r d4}

# Median Income of African American Household in 2020 
blackmedianincome <- get_acs(geography = "state",
                          variables = "B19013B_001",
                          year = 2020)
head(blackmedianincome)

```

## STEP 6: Extracting Data to CSV

Finally, let's export our data frame to a csv. Change the value names as needed to match your variables. This method works for both decennial and ACS datasets.

```{r d5}

write.csv(black2020, file = "black2020.csv", row.names = T)
write.csv(blackmedianincome, file = "blackmedianincome.csv", row.names = T)

```

Hopefully this will help you get started with Tidycensus. For more information or advice please consult this website: <https://walker-data.com/tidycensus/articles/basic-usage.html> Also please feel free to reach out to me on GroupMe or email: [myles.ndiritu\@morehouse.edu](mailto:myles.ndiritu@morehouse.edu){.email}
